Old people and low-income communities are most affected by COVID-19

### 2022-02-07 outputs/paper/paper.Rmd
add readme

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
library(kableExtra)
wd <- getwd()
wd
setwd("../..")
covid19_cases_cleaned <- read_csv("inputs/data/COVID19-cases-clean.csv")
head(covid19_cases_cleaned)
```

@@

To find out who was most affected, we obtained all of COVID-19 case reports from the City of Toronto's data portal. After some investigation, we found out that as the patient's age increases, the mortality and hospitalization rate also increases exponentially. The mortality rate of kids under 19 years old is 0.00007 and this number increases to 0.23 for people above 90 years old, by a factor of 3000. On top of that, low income or high density neighbourhood are hit the hardest during this pandemic. Among 140 different communities, the top 10 communities with the most COVID-19 cases have contributed 20% of Toronto's total COVID-19 cases.

Our findings have lots of implications, it identifies the groups of people and neighborhood that are most affected by COVID-19, this information will allow the government to better support them and come up with new proceudres that will
slow down the spread of COVID-19.

@@

```{r,include = FALSE}
# first find out which age group has the most reported covid 19 cases
covid_combined_by_age_group <-
  group_by(covid19_cases_cleaned, `Age Group`) %>%
  summarize(n())
names(covid_combined_by_age_group)[names(covid_combined_by_age_group) == 'n()'] <- 'Total Covid Cases'

fatal_covid_cases <- covid19_cases_cleaned %>% filter(`Outcome` == 'FATAL')
fatal_covid_case_combined_by_age_group <- fatal_covid_cases %>% 
  group_by(`Age Group`) %>% 
  summarize(n())
names(fatal_covid_case_combined_by_age_group)[names(fatal_covid_case_combined_by_age_group) == 'n()'] <- 'Fatal Cases'

total_case_and_fatal_case <- merge(covid_combined_by_age_group,fatal_covid_case_combined_by_age_group,by="Age Group")
total_case_and_fatal_case$Mortality_Rate <- with(total_case_and_fatal_case, round(`Fatal Cases` / `Total Covid Cases`, 5))

# ICU cases
cases_that_went_to_hospital <- covid19_cases_cleaned %>% filter(`Ever Hospitalized` == 'Yes');
cases_that_went_to_hospital_combined_by_age_group <- cases_that_went_to_hospital %>% 
  group_by(`Age Group`) %>%
  summarize(n())
names(cases_that_went_to_hospital_combined_by_age_group)[names(cases_that_went_to_hospital_combined_by_age_group) == 'n()'] <- 'Hospitalized'
total_case_fatal_hos <- merge(total_case_and_fatal_case, cases_that_went_to_hospital_combined_by_age_group, by = "Age Group")
total_case_fatal_hos$Hospital_Rate <- with(total_case_fatal_hos, round(`Hospitalized` / `Total Covid Cases`, 5))
total_case_fatal_hos
```

@@

```{r include = FALSE}
age_group_without_case_count <- subset( total_case_and_fatal_case, select = -c(`Total Covid Cases`, `Fatal Cases` ))
age_group_without_case_count$Type <- "Mortality Rate"
names(age_group_without_case_count)[names(age_group_without_case_count) == 'Mortality_Rate'] <- 'Rate'
another_frame <- subset(total_case_fatal_hos, select = -c(`Total Covid Cases`, `Fatal Cases`, `Hospitalized`, `Mortality_Rate`))
names(another_frame)[names(another_frame) == 'Hospital_Rate'] <- 'Rate'
another_frame$Type <- "Hospitalized Rate"
new <- rbind(age_group_without_case_count, another_frame)
new$`Age Group`[new$`Age Group`=="19 and younger"] <- "19 and younger"
new$`Age Group`[new$`Age Group`=="20 to 29 Years"] <- "20-29"
new$`Age Group`[new$`Age Group`=="30 to 39 Years"] <- "30-39"
new$`Age Group`[new$`Age Group`=="40 to 49 Years"] <- "40-49"
new$`Age Group`[new$`Age Group`=="50 to 59 Years"] <- "50-59"
new$`Age Group`[new$`Age Group`=="60 to 69 Years"] <- "60-69"
new$`Age Group`[new$`Age Group`=="70 to 79 Years"] <- "70-79"
new$`Age Group`[new$`Age Group`=="80 to 89 Years"] <- "80-89"
new$`Age Group`[new$`Age Group`=="90 and older"] <- "Above 90"
```

@@

```{r, include = FALSE}
covid_combined_by_postal_code <-
  group_by(covid19_cases_cleaned, `Neighbourhood Name`) %>%
  summarize(n())
names(covid_combined_by_postal_code)[names(covid_combined_by_postal_code) == 'n()'] <- 'n'
```

```{r, include = FALSE}
total_case_by_postal_sorted <- covid_combined_by_postal_code[order(-covid_combined_by_postal_code$n),]
total_case_by_postal_sorted
top_10 <- total_case_by_postal_sorted[1:10, ]
top_10
```

>>>>

```{r setup, include = FALSE}
# We load all necessary libraries
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
library(kableExtra)
library(here)
# Read the dataset, use here to navigate to working directory
covid19_cases_cleaned <- read_csv(here::here("inputs/data/COVID19-cases-clean.csv"))
head(covid19_cases_cleaned)
```

@@

To find out who was most affected, we obtained all of COVID-19 case reports from the City of Toronto's data portal [@opendatac]. After some investigation, we found out that as the patient's age increases, the mortality and hospitalization rate also increases exponentially. The mortality rate of kids under 19 years old is 0.00007 and this number increases to 0.23 for people above 90 years old, by a factor of 3000. On top of that, low income or high density neighbourhood are hit the hardest during this pandemic. Among 140 different communities, the top 10 communities with the most COVID-19 cases have contributed 20% of Toronto's total COVID-19 cases.

This report will first find out if there exists a correlation between age and the mortality rate of COVID-19. Then it will attempt to find out which community in Toronto is most affect by COVID-19. Our findings have lots of implications, it will help to identify the groups of people and neighborhood that are at high risk, this information will allow the government to better support them and come up with new procedures that will
slow down the spread of COVID-19.

@@

```{r,include = FALSE}
# first find out which age group has the most reported covid 19 cases
covid_combined_by_age_group <-
  group_by(covid19_cases_cleaned, `Age Group`) %>%
  summarize(n())
# Change column name from n() to Total Covid Cases
names(covid_combined_by_age_group)[names(covid_combined_by_age_group) == 'n()'] <- 'Total Covid Cases'

# Find out all fatal covid cases
fatal_covid_cases <- covid19_cases_cleaned %>% filter(`Outcome` == 'FATAL')
fatal_covid_case_combined_by_age_group <- fatal_covid_cases %>% 
  group_by(`Age Group`) %>% 
  summarize(n())
# Change the name
names(fatal_covid_case_combined_by_age_group)[names(fatal_covid_case_combined_by_age_group) == 'n()'] <- 'Fatal Cases'

# Combine the two data frame, now for each age group, we know the total covid case and the fatal ones
total_case_and_fatal_case <- merge(covid_combined_by_age_group,fatal_covid_case_combined_by_age_group,by="Age Group")
total_case_and_fatal_case$Mortality_Rate <- with(total_case_and_fatal_case, round(`Fatal Cases` / `Total Covid Cases`, 5))

# ICU cases
cases_that_went_to_hospital <- covid19_cases_cleaned %>% filter(`Ever Hospitalized` == 'Yes');
cases_that_went_to_hospital_combined_by_age_group <- cases_that_went_to_hospital %>% 
  group_by(`Age Group`) %>%
  summarize(n())

# Combine the two data frame and calculate the hospitalization rate by using Hospitalized / Total Covid Cases
names(cases_that_went_to_hospital_combined_by_age_group)[names(cases_that_went_to_hospital_combined_by_age_group) == 'n()'] <- 'Hospitalized'
total_case_fatal_hos <- merge(total_case_and_fatal_case, cases_that_went_to_hospital_combined_by_age_group, by = "Age Group")
total_case_fatal_hos$Hospital_Rate <- with(total_case_fatal_hos, round(`Hospitalized` / `Total Covid Cases`, 5))
total_case_fatal_hos
```

@@

```{r include = FALSE}
# Construct a new dataframe, it contains identical information, but is easier to plot
age_group_without_case_count <- subset( total_case_and_fatal_case, select = -c(`Total Covid Cases`, `Fatal Cases` ))
age_group_without_case_count$Type <- "Mortality Rate"
names(age_group_without_case_count)[names(age_group_without_case_count) == 'Mortality_Rate'] <- 'Rate'
another_frame <- subset(total_case_fatal_hos, select = -c(`Total Covid Cases`, `Fatal Cases`, `Hospitalized`, `Mortality_Rate`))
names(another_frame)[names(another_frame) == 'Hospital_Rate'] <- 'Rate'
another_frame$Type <- "Hospitalized Rate"
new <- rbind(age_group_without_case_count, another_frame)
new$`Age Group`[new$`Age Group`=="19 and younger"] <- "19 and younger"
new$`Age Group`[new$`Age Group`=="20 to 29 Years"] <- "20-29"
new$`Age Group`[new$`Age Group`=="30 to 39 Years"] <- "30-39"
new$`Age Group`[new$`Age Group`=="40 to 49 Years"] <- "40-49"
new$`Age Group`[new$`Age Group`=="50 to 59 Years"] <- "50-59"
new$`Age Group`[new$`Age Group`=="60 to 69 Years"] <- "60-69"
new$`Age Group`[new$`Age Group`=="70 to 79 Years"] <- "70-79"
new$`Age Group`[new$`Age Group`=="80 to 89 Years"] <- "80-89"
new$`Age Group`[new$`Age Group`=="90 and older"] <- "Above 90"
```

@@

```{r, include = FALSE}
# Group covid case by Neighbourhood Name
covid_combined_by_postal_code <-
  group_by(covid19_cases_cleaned, `Neighbourhood Name`) %>%
  summarize(n())
names(covid_combined_by_postal_code)[names(covid_combined_by_postal_code) == 'n()'] <- 'n'
```

```{r, include = FALSE}
# Sort by case counts, and get the first 10
total_case_by_postal_sorted <- covid_combined_by_postal_code[order(-covid_combined_by_postal_code$n),]
total_case_by_postal_sorted
top_10 <- total_case_by_postal_sorted[1:10, ]
top_10
```
